---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
options(warn=-1)
```

# interlacer <img src="man/figures/logo.svg" align="right" height="140" />

[![Project Status: WIP – Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)
[![R-CMD-check](https://github.com/khusmann/interlacer/actions/workflows/check-standard.yaml/badge.svg)](https://github.com/khusmann/interlacer/actions/workflows/check-standard.yaml)
[![codecov](https://codecov.io/gh/khusmann/interlacer/graph/badge.svg?token=R4WNWH5NXU)](https://codecov.io/gh/khusmann/interlacer)

When a value is missing in your data, sometimes you want to know *why* it is
missing. Many textual tabular data sources will encode missing reasons as
special values *interlaced* with the regular values in a column (e.g. `N/A`,
`REFUSED`, `-99`, etc.). Unfortunately, the missing reasons are
lost when these values are all converted into a single `NA` type. Working
with missing reasons in R traditionally requires loading variables 
as character vectors and doing a bunch of string
comparisons and type conversions to make sense of them.

`interlacer` provides functions that load variables from interlaced data sources
into a special `interlaced` column type that holds values and `NA`
reasons in separate *channels* of the same variable. In most contexts, you 
can treat `interlaced` columns as if they were regular values: if you take
the `mean` of an interlaced column, for example,
you get the mean of its values, without its missing reasons interfering in
the computation.

Unlike a regular column, however, the missing reasons are still
available. This means you can still filter data frames on variables
by specific missing reasons, or generate 
summary statistics with 
breakdowns by missing reason. In other words, you no longer have to constantly
manually include / exclude missing reasons in computations by filtering them
with awkward string comparisons or type conversions... everything just works!

In addition to the introduction in `vignette("interlacer")`
be sure to also check out:

- `vignette("coded-data")` for some recipies for working with coded data (e.g.
data produced by SPSS, SAS or Stata)

- `vignette("other-approaches")` for a deep dive into how interlacer's approach
compares to other approaches for representing and manipulating missing reasons
alongside data values

This library is currently in its experimental stages, so be aware that its
interface is likely to change in the future. In the meantime, please try it out
and
[let me know what you think](mailto:kdh38@psu.edu)!

## Installation

The easiest way to get `interlacer` is to install via devtools:

```{r, eval = FALSE}
install.packages("devtools") # If devtools is not already installed

devtools::install_github("khusmann/interlacer")
```

## Usage

To use interlacer, load it into your current R session:

```{r}
library(interlacer)
```

interlacer supports the following file formats with these `read_interlaced_*()`
functions, which extend the `readr::read_*()` family of functions:

* `read_interlaced_csv()`
* `read_interlaced_tsv()`
* `read_interlaced_csv2()`
* `read_interlaced_delim()`

As a quick demo, consider the following example file bundled with interlacer:

```{r}
library(dplyr, warn.conflicts = FALSE)
library(readr)

read_file(interlacer_example("colors.csv")) |>
  cat()
```

In this csv file, values are interlaced with three possible missing reasons:
`REFUSED`, `OMITTED`, and `N/A`.

With `readr`, loading these data would result in a data frame where all missing
reasons are replaced with `NA`:

```{r}
read_csv(
  interlacer_example("colors.csv"),
  na = c("REFUSED", "OMITTED", "N/A"),
  show_col_types = FALSE,
)
```

With interlacer, missing reasons are preserved:

```{r}
(ex <- read_interlaced_csv(
  interlacer_example("colors.csv"),
  na = c("REFUSED", "OMITTED", "N/A"),
  show_col_types = FALSE,
))
```

As you can see, in the printout above each column is defined by *two* types: a type
for values, and a type for missing reasons. The `age` column, for example, has type
`double` for its values, and type `character` for its missing reasons:

```{r}
ex$age
```

Computations automatically operate on values:

```{r}
mean(ex$age, na.rm=TRUE)
```

But the missing reasons are still there! To indicate a value should be treated
as a missing reason instead of a regular value, you can use the `na()` function.
The following, for example,
will filter the data set for all individuals that `REFUSED` to give their
favorite color:

```{r}
ex |>
  filter(favorite_color == na("REFUSED"))
```

And here's a pipeline that will compute a breakdown of the mean age of
respondents for each favorite color, with separate categories for each missing
reason:

```{r}
ex |>
  summarize(
    mean_age = mean(age, na.rm=T),
    n = n(),
    .by = favorite_color
  ) %>%
  arrange(favorite_color)
```

But this just scratches the surface of what can be done with interlacer...
check out `vignette("interlacer")` for a more complete overview!

## Known Issues

1. Interlaced vectors cannot be constructed with `c()`

Due to a [fundamental limitation of R](https://vctrs.r-lib.org/#motivation),
interlaced vectors cannot be constructed with `c()`. For example, the following
will return a regular vector:

```{r}
c(5, 6, na("OMITTED"))
```

Instead, use `vctrs::vec_c()` (reexported by `interlacer`):

```{r}
vec_c(5, 6, na("OMITTED"))
```

2. `ifelse()` does not promote interlaced types

For similar reasons as the above issue, using `base::ifelse()` with `interlaced`
columns will convert them into regular vectors:

```{r}
ex |>
  mutate(
    favorite_color = ifelse(age < 18, na("REDACTED"), favorite_color)
  )
```

Instead, use `dplyr::if_else()`:

```{r}
ex |>
  mutate(
    favorite_color = if_else(
      age < 18,
      na("REDACTED_UNDERAGE"),
      favorite_color,
      missing = na("REDACTED_MISSING_AGE")
    )
  )
```

3. Performance with large data sets

You may notice that on large datasets `interlacer` runs significantly slower
than `readr` / `vroom`. Although `interlacer` uses `vroom` under the hood to load 
delimited data, it is not able to take advantage of many of its optimizations
because `vroom` does not
[does not currently support](https://github.com/tidyverse/vroom/issues/532)
column-level missing values.  As soon as `vroom` supports column-level
missing values, I will be able to remedy this!

## Related work

`interlacer` was inspired by the [`haven`](https://haven.tidyverse.org/),
[`labelled`](https://larmarange.github.io/labelled/), and
[`declared`](https://dusadrian.github.io/declared/) packages. These packages
provide similar functionality to `interlacer`, but are more focused on
providing compatibility
with missing reason data imported from SPSS, SAS, and Stata. `interlacer`, by
contrast, aims to be more generic: In addition to having the ability to
model SPSS, SAS, and Stata
missing value labels, it allows you to compose *any* two R types into an
`interlaced` vector that can be easily manipulated in tidy pipelines.
For a more detailed discussion, see `vignette("other-approaches")`.

Future versions will include conversion functions to and from the types provided
by these other packages!

## Acknowledgements

The development of this software was supported, in whole or in part, by the Institute of Education Sciences, U.S. Department of Education, through Grant R305A170047 to The Pennsylvania State University. The opinions expressed are those of the authors and do not represent the views of the Institute or the U.S. Department of Education.
