---
title: "Coded Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Coded Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In addition to interlacing values and missing reasons, many statistical software
packages will store categorical values and missing reasons as alphanumeric
codes. These codes are often chosen so that numeric comparisons or casts can be
used to determine if a value represents a real value or missing reason.
Like 8-character variable name limits, this practice comes from a
historical need to save digital storage space even if it made analyses less
readable and more error-prone.

Even though storage is cheap these days, coded formats continue to be the
standard format used
by statistical software packages like SPSS, SAS, and Stata. This article
will describe these common coding schemes and how they can be decoded and
deinterlaced to make them easier to work with in R.

## Numeric codes with negative missing reasons (SPSS)

It's extremely common to find data sources that encode all categorical responses
as numeric values, with negative values representing missing reason codes. SPSS
is one such example. Here's an SPSS-formatted version of the `colors.csv`
example:

```{r}
library(readr)
library(interlacer)

read_file(
  interlacer_example("colors_coded.csv")
) |>
  cat()
```

Where missing reasons are:

> -99: N/A
>
> -98: REFUSED
>
> -97: OMITTED

And colors are coded:

> 1: BLUE
>
> 2: RED
>
> 3: YELLOW

This format gives you the ability to load everything as a numeric type:

```{r}
(df_coded <- read_csv(
  interlacer_example("colors_coded.csv"),
  col_types = "n"
))
```

To test if a value is a missing code, you can check if it's less than 0:

```{r}
library(dplyr, warn.conflicts = FALSE)

df_coded |>
  mutate(
    favorite_color_missing = if_else(favorite_color < 0, favorite_color, NA),
    age = if_else(age > 0, age, NA)
  ) |>
  summarize(
    mean_age = mean(age, na.rm=T),
    n = n(),
    .by = favorite_color_missing
  )
```

The downsides of this approach are twofold: 1) all of your values and 
missing reasons become codes you have to remember and 2) it's really easy to
make mistakes.

What sort of mistakes? Well, because everything is numeric, there's nothing
stopping us from treating missing reason codes as if they are regular values...
If you forget to remove your missing reason codes, R will still happily
compute aggregations using the negative numbers!

```{r}
df_coded |>
  mutate(
    favorite_color_missing = if_else(favorite_color < 0, favorite_color, NA),
#    age = if_else(age > 0, age, NA)
  ) |>
  summarize(
    mean_age = mean(age, na.rm=T),
    n = n(),
    .by = favorite_color_missing
  )
```

In fact, ANY math you do without filtering for missing codes potentially ruins
the integrity of your data:

```{r}
# This will add 1 to the age values, but ALSO add one to all of the missing
# reason codes, resulting in corrupted data!
df_coded |>
  mutate(
    age_next_year = age + 1,
    .after = person_id
  )

# This will give you your intended result, but it's easy to forget
df_coded |>
  mutate(
    age_next_year = if_else(age < 0, age, age + 1),
    .after = person_id
  )
```

Have you ever thought you had a significant result, only to find that it's
only because there are some stray missing reason codes still interlaced with
your values? It's a bad time.

You're much better off loading these formats with interlacer, then converting
the codes into labelled factor levels:

```{r}
library(forcats)

(df_decoded_deinterlaced <- read_interlaced_csv(
  interlacer_example("colors_coded.csv"),
  na = c("-99", "-98", "-97")
) |>
  mutate(
    across(
      missing_cols(),
      \(x) fct_recode(x,
        `N/A` = "-99",
        REFUSED = "-98",
        OMITTED = "-97",
      )
    ),
    favorite_color = fct_recode(
      as.character(favorite_color),
      BLUE = "1",
      RED = "2",
      YELLOW = "3",
    )
  ))
```

Now aggregations won't mix up values and missing codes, and you won't have to
keep cross-referencing your codebook to know what values mean:

```{r}
df_decoded_deinterlaced |>
  summarize(
    mean_age = mean(age, na.rm=T),
    n = n(),
    .by = .favorite_color.
  )
```

Other operations work with similar ease:

```{r}
df_decoded_deinterlaced |>
  mutate(
    age_next_year = age + 1,
    .after = person_id
  ) |>
  coalesce_channels(default_reason = "AGE_UNAVAILABLE")
```



## Numeric codes with character missing reasons (SAS, Stata)

Like SPSS, SAS and Stata will encode factor levels as numeric values, but
instead of representing missing reasons as negative codes, they are given
character codes:

```{r}
read_file(
  interlacer_example("colors_coded_char.csv")
) |>
  cat()
```

Here, the same value codes are used as the previous example, except the missing
reasons are coded as follows:

> ".": N/A
>
> ".a": REFUSED
>
> ".b": OMITTED


To handle these missing reasons without interlacer, columns must be loaded as
character vectors:

```{r}
(df_coded_char <- read_csv(
  interlacer_example("colors_coded_char.csv"),
  col_types = "c"
))
```

To test if a value is missing, they can be cast to numeric types. If the cast fails,
you know it's a missing code. If it is successful, you know it's a coded value.

```{r}
df_coded_char |>
  mutate(
    favorite_color_missing = if_else(
      is.na(as.numeric(favorite_color)),
      favorite_color,
      NA
    ),
    age = if_else(!is.na(as.numeric(age)), as.numeric(age), NA)
  ) |>
  summarize(
    mean_age = mean(age, na.rm=T),
    n = n(),
    .by = favorite_color_missing
  )
```

Although the character missing codes help prevent us from mistakenly including
missing codes in value aggregations, having to cast our columns to numeric
all the time to check for missingness is hardly ergonomic, and generates
annoying warnings. Like before, it's easier to import with interlacer and
decode the values and missing reasons:

```{r}
read_interlaced_csv(
  interlacer_example("colors_coded_char.csv"),
  na = c(".", ".a", ".b")
) |>
  mutate(
    across(
      missing_cols(),
      \(x) fct_recode(x,
        `N/A` = ".",
        REFUSED = ".a",
        OMITTED = ".b",
      )
    ),
    favorite_color = fct_recode(
      as.character(favorite_color),
      BLUE = "1",
      RED = "2",
      YELLOW = "3",
    )
  )
```

## Encoding a decoded & deinterlaced data frame.

Re-coding and re-interlacing a data frame is easily done as follows:

```{r, eval = FALSE}
df_decoded_deinterlaced |>
  mutate(
    across(
      missing_cols(),
      \(x) fct_recode(x,
        `-99` = "N/A",
        `-98` = "REFUSED",
        `-97` = "OMITTED"
      )
    ),
    favorite_color = fct_recode(
      favorite_color,
      `1` = "BLUE",
      `2` = "RED",
      `3` = "YELLOW"
    )
  ) |>
  write_interlaced_csv("output.csv")
```

## haven

The [haven](https://haven.tidyverse.org/) package has functions for loading
native SPSS, SAS, and Stata native file formats into
special data frames that use column attributes and special values to keep track
of interlaces values and missing reasons. For a complete discussion of how this
compares to interlacer's approach, see `vignette("other-approaches")`.

Future versions of interlacer could have the ability to convert haven data
frames to and from deinterlaced data frames, but I want to gauge interest for
this feature before I invest the time to implement it. If this is a a feature
you'd use, [please let me know](mailto:kdh38@psu.edu)!
