---
title: "Interlaced Column Types"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interlaced Column Types}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Like the `readr::read_*()` family of functions, `read_interlaced_*()` will
automatically guess column types by default:

```{r}
library(interlacer)

(read_interlaced_csv(
  interlacer_example("colors.csv"),
  na = c("REFUSED", "OMITTED", "N/A"),
))
```

As with readr, these column type guess can be overridden using the `col_types`
parameter with readr's `readr::col_*()` column specifiers:

```{r}
library(readr)

(read_interlaced_csv(
  interlacer_example("colors.csv"),
  na = c("REFUSED", "OMITTED", "N/A"),
  col_types = cols(
    person_id = col_integer(),
    age = col_number(),
    favorite_color = col_factor(levels = c("BLUE", "RED", "YELLOW", "GREEN"))
  )
))
```

## Interlaced Column Types

In addition to the standard `readr::col_*` column specification types, 
interlacer provides *interlaced* column types that enable you to specify missing
reasons at the column level.

This is useful when you have missing reasons that only apply to particular items
as opposed to the file as a whole. For example, say we had a measure with the
following two items:

> 1. What is your current stress level?
> a. Low
> b. Moderate
> c. High
> d. I don't know
> e. I don't understand the question
> 
> 2. How well do you feel you manage your time and responsibilities today?
> a. Poorly
> b. Fairly well
> c. Well
> d. Very well
> e. Does not apply (Today was a vacation day)
> f. Does not apply (Other reason)

As you can see, both items have two selection choices that should be mapped to
missing reasons. To specify missing reasons at the variable level, the 
`icol_*()` family
of column specification types can be used. These extend all of readr's
`col_*()` column types by adding a parameter for specifying missing values
unique to that particular variable:

```{r}
(df_stress <- read_interlaced_csv(
  interlacer_example("stress.csv"),
  col_types = cols(
    person_id = col_integer(),
    current_stress = icol_factor(
      levels = c("LOW", "MODERATE", "HIGH"),
      na = c("DONT_KNOW", "DONT_UNDERSTAND")
    ),
    time_management = icol_factor(
      levels = c("POORLY", "FAIRLY_WELL", "WELL", "VERY_WELL"),
      na = c("NA_VACATION", "NA_OTHER")
    )
  ),
  na = c(
    "REFUSED",
    "OMITTED",
    "N/A"
  )
))
```

## Setting missing reason types with `na_col_types`

interlacer will automatically guess the type of missing values as well. For
example, when we load a dataset that uses numeric codes for missing reasons,
it results in numeric missing reason channels.

```{r}
read_interlaced_csv(
  interlacer_example("colors_coded.csv"),
  na = c(-99, -98, -97)
)
```

The type of missing reason channels can be explicitly set via the `na_col_types`
parameter, which accepts a `cols()` specification. The following will read
each missing reason channel as a character column:

```{r}
read_interlaced_csv(
  interlacer_example("colors_coded.csv"),
  na = c(-99, -98, -97),
  na_col_types = cols(
    person_id = col_character(),
    age = col_character(),
    favorite_color = col_character(),
  )
)
```

`col_skip()` can be used to omit the missing reason channel loading the column
as a regular type instead of an `interlaced` one. `.default` also can be used
to set the default column type:

```{r}
read_interlaced_csv(
  interlacer_example("colors_coded.csv"),
  na = c(-99, -98, -97),
  na_col_types = cols(
    person_id = col_skip(),
    .default = col_integer(),
  )
)
```

## Next steps

In this vignette we covered how the column types for values and missing reasons
can be explicitly specified using collectors. We also illustrated how
column-level missing values can be specified using the `icol_*` collector types.

In the final example, we used an example data set with coded values and missing
reasons. Coded values are especially common in data sets produced by SPSS, SAS,
and Stata. For some recipes for working with coded data like this, check
out the next vignette, `vignette("coded-data")`.
