---
title: "Interlaced Column Types"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interlaced Column Types}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Like the `readr::read_*()` family of functions, `read_interlaced_*()` will
automatically guess column types by default:

```{r}
library(interlacer)

(read_interlaced_csv(
  interlacer_example("colors.csv"),
  na = c("REFUSED", "OMITTED", "N/A"),
))
```

As with readr, these column type guess can be overridden using the `col_types`,
using readr's `readr::col_*()` column specifiers:

```{r}
library(readr)

(read_interlaced_csv(
  interlacer_example("colors.csv"),
  na = c("REFUSED", "OMITTED", "N/A"),
  col_types = cols(
    person_id = col_integer(),
    age = col_number(),
    favorite_color = col_factor(levels = c("BLUE", "RED", "YELLOW", "GREEN"))
  )
))
```

## Interlaced Column Types

In addition to the standard `readr::col_*` column specification types, 
interlacer provides *interlaced* column types that enable you to specify missing
reasons at the column level.

This is useful when you have missing reasons that only apply to particular items
as opposed to the file as a whole. For example, say we had a measure with the
following two items:

> 1. What is your current stress level?
> a. Low
> b. Moderate
> c. High
> d. I don't know
> e. I don't understand the question
> 
> 2. How well do you feel you manage your time and responsibilities today?
> a. Poorly
> b. Fairly well
> c. Well
> d. Very well
> e. Does not apply (Today was a vacation day)
> f. Does not apply (Other reason)

As you can see, both items have two selection choices that should be mapped to
missing reasons. To specify missing reasons at the variable level, the 
`icol_*()` family
of column specification types can be used. These extend all of readr's
`col_*()` column types by adding a parameter for specifying missing values
unique to that particular variable:

```{r}
(df_stress <- read_interlaced_csv(
  interlacer_example("stress.csv"),
  col_types = cols(
    person_id = col_integer(),
    current_stress = icol_factor(
      levels = c("LOW", "MODERATE", "HIGH"),
      na = c("DONT_KNOW", "DONT_UNDERSTAND")
    ),
    time_management = icol_factor(
      levels = c("POORLY", "FAIRLY_WELL", "WELL", "VERY_WELL"),
      na = c("NA_VACATION", "NA_OTHER")
    )
  ),
  na = c(
    "REFUSED",
    "OMITTED",
    "N/A"
  )
))

```

As you can see, the `icol_factor()` column spec works just like
`readr::col_factor()`, but additionally accepts an `na` argument for specifying
missing values at the variable level. When you specify missing
reasons at the variable-level like this, the available levels in the resulting
missing reason column correctly show only the possible missing reasons for
that variable:

```{r}
levels(df_stress$.person_id.)
levels(df_stress$.current_stress.)
levels(df_stress$.time_management.)
```

For comparison, if we loaded all of these variable-level missing reasons as
file-level level missing reasons, each variable would have ALL missing
reasons as possible levels, even if they didn't apply to that particular
variable:

```{r}
df_stress_file <- read_interlaced_csv(
  interlacer_example("stress.csv"),
  na = c(
    "REFUSED",
    "OMITTED",
    "N/A",
    "DONT_KNOW",
    "DONT_UNDERSTAND",
    "NA_VACATION",
    "NA_OTHER"
  )
)

levels(df_stress_file$.person_id.)
levels(df_stress_file$.current_stress.)
levels(df_stress_file$.time_management.)
```
