---
title: "Extended Column Types"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extended Column Types}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Like the `readr::read_*()` family of functions, `read_interlaced_*()` will
automatically guess column types by default:

```{r}
library(interlacer, warn.conflicts = FALSE)

(read_interlaced_csv(
  interlacer_example("colors.csv"),
  na = c("REFUSED", "OMITTED", "N/A"),
  show_col_types = FALSE
))
```

As with readr, these column type guess can be overridden using the `col_types`
parameter with a `readr::cols()` column specification:

```{r}
library(readr)

(read_interlaced_csv(
  interlacer_example("colors.csv"),
  col_types = cols(
    person_id = col_integer(),
    age = col_number(),
    favorite_color = col_factor(levels = c("BLUE", "RED", "YELLOW", "GREEN"))
  ),
  na = c("REFUSED", "OMITTED", "N/A")
))
```

## `x_cols`: extended `cols` specifications

When you need more fine-grained control over value and missing reason channel
types, you can use an `x_cols()` specification, an extension of readr's 
`readr::cols()` system. With `x_cols()`, you can control both the value channel
and na channel types in the columns of the resulting data frame. 

This is useful when you have missing reasons that only apply to particular items
as opposed to the file as a whole. For example, say we had a measure with the
following two items:

1. What is your current stress level?

> a. Low
> b. Moderate
> c. High
> d. I don't know
> e. I don't understand the question

2. How well do you feel you manage your time and responsibilities today?

> a. Poorly
> b. Fairly well
> c. Well
> d. Very well
> e. Does not apply (Today was a vacation day)
> f. Does not apply (Other reason)

As you can see, both items have two selection choices that should be mapped to
missing reasons. These can be specified with the `x_cols()` as follows:

```{r}
(df_stress <- read_interlaced_csv(
  interlacer_example("stress.csv"),
  col_types = x_cols(
    person_id = x_col(
      v_col_integer(),
      na_col_none()
    ),
    current_stress = x_col(
      v_col_factor(levels = c("LOW", "MODERATE", "HIGH")),
      na_col_factor("DONT_KNOW", "DONT_UNDERSTAND")
    ),
    time_management = x_col(
      v_col_factor(levels = c("POORLY", "FAIRLY_WELL", "WELL", "VERY_WELL")),
      na_col_factor("NA_VACATION", "NA_OTHER")
    )
  )
))
```

Like readr's `readr::cols()` function, each named `x_cols()` describes a column
in the resulting data frame. Value and missing reason channel types are 
declared via calls to `v_col_*()` and `na_col_*()` respectively, which are
assembled by `x_col()`.

`v_col_*()` types mirror readr's `readr::col_*` column "collectors". So
`v_col_double()` is equivalent to `readr::col_double()`, `v_col_character()` is
equivalent to `readr::col_character()`, etc. See vroom's documentation for a
[list of available column types](https://vroom.r-lib.org/articles/vroom.html#column-types).

`na_col_*()` collectors allow you to declare missing reason channel type of the
loaded column, and the values that should be interpreted as missing reasons.
Currently, there are five options:

1. `na_col_default()`: Use the collector defined by the `na = ` argument
in the `read_interlaced_*()` function

2. `na_col_none()`: Load the column without a missing reason channel.

3. `na_col_factor()`: Use a factor missing reason channel. Character arguments
passed form the levels of the factor.
(e.g. `na_col_factor("REFUSED", "OMITTED", "N/A")`)

4. `na_col_integer()`: Use an integer na channel. Numeric arguments passed are
the values to be interpreted as missing values. (e.g.
`na_col_integer(-99, -98, -97))`)

5. `na_col_cfactor()`: Use a `cfactor()` na channel. (`cfactor()` types will
be covered in the next vignette,`vignette("coded-data")`)

The following example shows some of these collectors in action. In this example
we use a coded version of the `colors.csv` example data, to demonstrate integer
missing reason types:

```{r}
read_interlaced_csv(
  interlacer_example("colors_coded.csv"),
  col_types = x_cols(
    person_id = x_col(v_col_integer(), na_col_none()),
    age = x_col(v_col_double(), na_col_integer(-99, -98, -97)),
    favorite_color = x_col(v_col_integer(), na_col_integer(-99, -98, -97))
  )
)
```

## Shortcuts

### Default collector types

Like readr's `cols()` function, the `x_cols()` function accepts a `.default`
argument that specifies a default value collector. The `na = ` argument is
similarly used to specify a default missing reason collector to be used when
no `na_col_*()` is specified, or when it is set to `na_col_default()`.

By taking advantage of these defaults, the specification in the last example
could have been equivalently written as:

```{r}
read_interlaced_csv(
  interlacer_example("colors_coded.csv"),
  col_types = x_cols(
    .default = v_col_integer(),
    person_id = x_col(v_col_integer(), na_col_none()),
    age = v_col_double(),
  ),
  na = na_col_integer(-99, -98, -97)
)
```

### Concise value and missing reason specifications

Like readr, value collectors can be specified using characters. For example,
instead of `v_col_integer()`, you can use `"i"`. See vroom's documentation for a
[complete list of these shortcuts](https://vroom.r-lib.org/articles/vroom.html#column-types).

Similarly, missing reason collectors can be specified by providing a vector of
the missing values; the collector type is inferred via the type of the vector.
The conversions are as follows:

- `na_col_none()`: `NULL`

- `na_col_factor()`: A character vector, e.g. `c("REFUSED", "OMITTED", "N/A")`

- `na_col_integer()`: A numeric vector, e.g. `c(-99, -98, -97))`

- `na_col_cfactor()`: A named character or numeric vector, e.g.
``c(REFUSED = -99, OMITTED = -98, `N/A` = -97)``

Using these shortcuts, the previous example could have equivalently been written
in more compact form as follows:

```{r}
read_interlaced_csv(
  interlacer_example("colors_coded.csv"),
  col_types = x_cols(
    .default = "i",
    person_id = x_col("i", NULL),
    age = "d",
  ),
  na = c(-99, -98, -97)
)
```

## Next steps

In this vignette we covered how the column types for values and missing reasons
can be explicitly specified using collectors. We also illustrated how
column-level missing values can be specified by creating extended column type
specifications using `x_cols()`.

In the final examples, we used an example data set with coded values and missing
reasons. Coded values are especially common in data sets produced by SPSS, SAS,
and Stata. interlacer provides a special column type to make working with this
sort of data easier: the `cfactor()` type. This will be covered in next
vignette, `vignette("coded-data")`.
