---
title: "Extended Column Types"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extended Column Types}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Like the `readr::read_*()` family of functions, `read_interlaced_*()` will
automatically guess column types by default:

```{r}
library(interlacer, warn.conflicts = FALSE)

(read_interlaced_csv(
  interlacer_example("colors.csv"),
  na = c("REFUSED", "OMITTED", "N/A"),
  show_col_types = FALSE
))
```

As with readr, these column type guess can be overridden using the `col_types`
parameter with readr's `readr::col_*()` column specifiers:

```{r}
library(readr)

(read_interlaced_csv(
  interlacer_example("colors.csv"),
  na = c("REFUSED", "OMITTED", "N/A"),
  col_types = cols(
    person_id = col_integer(),
    age = col_number(),
    favorite_color = col_factor(levels = c("BLUE", "RED", "YELLOW", "GREEN"))
  )
))
```

## `NA` collector types

In addition to the standard `readr::col_*` column specification types, 
interlacer provides the ability to specify missing
reasons at the column level, using the `na` parameter.

This is useful when you have missing reasons that only apply to particular items
as opposed to the file as a whole. For example, say we had a measure with the
following two items:

1. What is your current stress level?

> a. Low
> b. Moderate
> c. High
> d. I don't know
> e. I don't understand the question

2. How well do you feel you manage your time and responsibilities today?

> a. Poorly
> b. Fairly well
> c. Well
> d. Very well
> e. Does not apply (Today was a vacation day)
> f. Does not apply (Other reason)

As you can see, both items have two selection choices that should be mapped to
missing reasons. These can be specified with the `na_cols()` function, which
works similarly to readr's `cols()` function:

```{r}
usual_na_vals <- c("REFUSED", "OMITTED", "N/A")

(df_stress <- read_interlaced_csv(
  interlacer_example("stress.csv"),
  col_types = x_cols(
    person_id = x_col(v_col_integer(), na_col_none()),
    current_stress = x_col(
      v_col_factor(levels = c("LOW", "MODERATE", "HIGH")),
      c(usual_na_vals, "DONT_KNOW", "DONT_UNDERSTAND")
    ),
    time_management = x_col(
      v_col_factor(levels = c("POORLY", "FAIRLY_WELL", "WELL", "VERY_WELL")),
      c(usual_na_vals, "NA_VACATION", "NA_OTHER")
    )
  ),
  na = usual_na_vals,
))
```


Setting na type to `NULL` indicates the column should be loaded
as a regular type instead of an `interlaced` one. The following
will load `person_id` as a regular, non-interlaced type:

```{r}
read_interlaced_csv(
  interlacer_example("colors_coded.csv"),
  col_types = x_cols(
    person_id = x_col(v_col_guess(), na_col_none())
  ),
  na = c(-99, -98, -97),
  show_col_types = FALSE
)
```

## Next steps

In this vignette we covered how the column types for values and missing reasons
can be explicitly specified using collectors. We also illustrated how
column-level missing values can be specified by creating a missing channel
specification using `na_cols()`.

In the final example, we used an example data set with coded values and missing
reasons. Coded values are especially common in data sets produced by SPSS, SAS,
and Stata. For some recipes for working with coded data like this, check
out the next vignette, `vignette("coded-data")`.
