---
title: "Introduction to interlacer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to interlacer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(readr.show_col_types = FALSE)
```

In many datasets, reasons for missing values are interlaced with data as special
values or codes. For example, consider the following CSV:

```{r}
library(readr)
library(interlacer, warn.conflicts = FALSE)

read_file(interlacer_example("colors.csv")) |>
  cat()
```

This csv data source has three variables: `person_id` and `age`,
both numeric variables and `favorite_color`, a character or factor variable.
Interlaced in their values are three possible missing reasons: `REFUSED`,
`OMITTED`, and `N/A`.

To load the values of this data source, it is an easy call to the venerable
`readr::read_csv()`:

```{r}
(df_simple <- read_csv(
  interlacer_example("colors.csv"),
  na = c("REFUSED", "OMITTED", "N/A")
))
```

As you can see, the data were loaded into a dataframe with three columns,
and all of the missing reasons were replaced with `NA` values.

## Aggregations with missing reasons

Now, if we were only interested in the *values* of our source data, this
functionality is all we need. But what if we wanted to know *why* some values
were `NA`? Although that information was encoded in our source data, it was
lost when all of the missing reasons were converted into `NA` values.

For example, consider the `favorite_color` column. How many respondents 
`REFUSED` to give their favorite color? How many people just `OMITTED` their 
answer? Was the question `N/A` for some respondents (e.g. wasn't on their
survey form)? What was the mean respondent age for each of these groups?

Our current dataframe only gets us part way:

```{r}
library(dplyr, warn.conflicts = FALSE)

df_simple |>
  summarize(
    mean_age = mean(age, na.rm = TRUE),
    n = n(),
    .by = favorite_color
  ) |>
  arrange(favorite_color)
```

As you can see, because we converted all our missing reasons into a single `NA`,
we can only answer these questions about missingness in general, rather than
work with the specific reasons stored in our source data.

Unfortunately, if we try load our data with the missing reasons intact, we lose
something else: the type information of the values.

```{r}
(df_with_missing <- read_csv(
  interlacer_example("colors.csv"),
  col_types = cols(.default = "c")
))
```

Now we have access to our missing reasons, but all the columns are character
vectors. This means that in order to do anything with our values, we always
have to filter out the missing reasons, and cast the remaining values to our 
desired type:

```{r}
reasons <- c("REFUSED", "OMITTED", "N/A")

df_with_missing |>
  mutate(
    age_values = as.numeric(if_else(age %in% reasons, NA, age)),
  ) |>
  summarize(
    mean_age = mean(age_values, na.rm = TRUE),
    n = n(),
    .by = favorite_color
  ) |>
  arrange(favorite_color)
```

This gives us the information we want, but it is cumbersome. Notice how there's
no distinction between favorite color values and missing reasons! Things start
to get
really complex when different columns have different sets of possible missing
reasons. It means you have to do a lot of type conversion
gymnastics to switch between value types and missing types.

### The interlacer approach

interlacer was built based on the insight that everything becomes much more
tidy, simple, and expressive when we explicitly work with values and missing
reasons as separate *channels* of the same variable. interlacer
introduces a new `interlaced` column type that facilitates this. 
The `read_interlaced_*` functions in interlacer import data with this new
column type.

```{r}
(df <- read_interlaced_csv(
  interlacer_example("colors.csv"),
  na = c("REFUSED", "OMITTED", "N/A")
))
```

As you can see by the column headers, each column loaded is composed of two
channels: a value channel,
and missing reason channel. Each channel can have its own type. The `age`
column, for example, has `double` values and `factor` missing reasons:

```{r}
df$age
```

These channels can be explicitly accessed using the `value_channel()` and
`na_channel()` helper functions:

```{r}
value_channel(df$age)
na_channel(df$age)
```

These helpers are rarely needed, however, because computations automatically
operate on an `interlaced` column's value channel,
and ignore the missing reasons channel. The following will compute the mean
age, without the missing reasons interfering:

```{r}
mean(df$age, na.rm = TRUE)
```

(We could have equivalently used the `value_channel()` helper to achieve the
same result, albeit with more verbosity):

```{r}
mean(value_channel(df$age), na.rm = TRUE)
```

Although missing reasons are excluded in computations, they still treated as
unique values. This means that when we group by age we get a breakdown by 
the unique missing reasons, rather than being lumped into a single `NA`:

```{r}
df |>
  summarize(
    mean_age = mean(age, na.rm = TRUE),
    n = n(),
    .by = favorite_color
  ) |>
  arrange(favorite_color)
```

As you can see, we can generate the same report as we did before, but without
needing to do any type gymnastics! Also, the values are neatly distinguished
from the missing reasons.

## Filtering based on missing reasons

`interlaced` columns are also helpful when creating
samples with inclusion / exclusion criteria based on missing reasons. For
example, using our example data, say we wanted to create a sample of respondents
that `REFUSED` to give their age. To indicate that a value should be interpreted
as a missing reason, you can use the `na()` function with the value:

```{r}
df |>
  filter(age == na("REFUSED"))

# na_channel() can also be used to get an equivalent result:
df |>
  filter(na_channel(age) == "REFUSED")
```

How about people who `REFUSED` to report their age AND favorite color?

```{r}
df |>
  filter(age == na("REFUSED") & favorite_color == na("REFUSED"))
```

It's also possible to combine value conditions with missing reason
conditions. For example, this will select everyone who `REFUSED` to give
their favorite color, and was over 20 years old:

```{r}
df |>
  filter(age > 20 & favorite_color == na("REFUSED"))
```

## Mutations

As you might expect, the `na()` function can be used with values in mutations. The
following pipeline will replace the favorite color of respondents with the missing
value `"REDACTED"`

```{r}
df |>
  mutate(
    favorite_color = na("REDACTED")
  )
```

Conditionals also work exactly as you would expect in mutations.
The following will replace the favorite color of respondents with `age < 18`
with the missing reason `"REDACTED_UNDERAGE"`. Respondents missing an age will
be replaced with `"REDACTED_MISSING_AGE"`

```{r}
df |>
  mutate(
    favorite_color = if_else(
      age < 18,
      na("REDACTED_UNDERAGE"),
      favorite_color,
      missing = na("REDACTED_MISSING_AGE")
    )
  )
```

The following mutation will create a new column called `person_type`
that will be `"CHILD"` when `age < 18`, `"ADULT"` when `age >= 18`, and missing
with reason `"AGE_UNAVAILABLE"` when `age` is missing:

```{r}
df |>
  mutate(
    person_type = if_else(
      age < 18,
      "CHILD",
      "ADULT",
      missing = na("AGE_UNAVAILABLE")
    ),
  )
```

Important note: You must use `dplyr::if_else()` with `interlaced` vectors
instead of R's `base::ifelse()` function, because the base function strips out
the missing reason channel due to a
[fundamental limitation in base R](https://vctrs.r-lib.org/#motivation).

## Multichannel comparison

Comparison operators (e.g. `==` and `!=`) compare value channels, and
return `NA` whenever missing values are involved:

```{r}
5 == NA
5 == na("reason")
```

This means that missing reasons cannot be directly compared:

```{r}
na("reason1") == na("reason2")
```

To simultaneously compare values and missing reason channels, interlacer
provides the multichannel comparison operators `%==%` and `%!=%`. These
operators always return boolean vectors, and never return missing values.
With multichannel comparison, two values are considered equal iff:

- they are both values, and the values are equal OR
- they are both missing values, and their reasons are the same OR
- they are both missing values, and both lack missing reasons

This way, if we convert the previous example to use multichannel comparison,
we get:

```{r}
na("reason1") %==% na("reason2")
```

Similarly, comparisons with values return booleans rather than `NA`,

```{r}
5 %==% NA
5 %==% na("reason")
```

And, as a consequence, regular `NA` values are equivalent to interlaced values
that do not have a missing reason:

```{r}
NA %==% na()

# Alternatively this can be considered an `NA` missing reason:
NA %==% na(NA)
```

Missing values that are missing their missing reasons are called "empty values",
and discussed in the next section.

## Empty values

Empty values are missing values that lack missing reasons. These values can
arise when missing actions are not
specified. For example, in the example, empty values are created because
the `missing = ` argument in the `if_else()` statement is not specified:

```{r}
df |>
  mutate(
    favorite_color = if_else(
      age < 18,
      na("REDACTED_UNDERAGE"),
      favorite_color,
    )
  )
```

As you can see, empty values are displayed as `<<NA>>`.

Empty values can be detected by using the multichannel comparsion operator:

```{r}
df |>
  mutate(
    favorite_color = if_else(
      age < 18,
      na("REDACTED_UNDERAGE"),
      favorite_color,
    )
  ) |>
  filter(favorite_color %==% na())
```

(Note that regular equality does not work, because it would return `NA` for all
the comparisons!)

Empty values often occur as the result of joins, because the `dplyr::*_join()`
family of functions do not have a `missing = ` parameter,
like `dplyr::if_else()` does. For example,
say we had the following data frame we wanted to join to our sample:

```{r}
conditions <- tribble(
  ~person_id, ~condition,
  1, "TREATMENT",
  2, "CONTROL",
  3, na("TECHNICAL_ERROR"),
  6, "CONTROL",
  8, "TREATMENT",
)
```

Because we're missing condition information for some of the respondents,
these show up as empty values when we join the data frame to our sample:

```{r}
df |>
  left_join(conditions, by = join_by(person_id))
```

We can remedy this by replacing these empty values after the join:

```{r}
df |>
  left_join(conditions, by = join_by(person_id)) |>
  mutate(
    condition = if_else(condition %==% na(), na("LEFT_STUDY"), condition),
  )
```

## Writing interlaced files

After you've made made changes to your data, you probably want to save them.
interlacer provides the `write_interlaced_*` family of functions for this:

```{r, eval = FALSE}
write_interlaced_csv(df, "interlaced_output.csv")
```

This will combine the value and missing reasons into interlaced character
columns, and write the result as a csv. Alternatively, if you want to
re-interlace the columns without writing to a file for more control in the
writing process, you can use `flatten_channels()`:

```{r}
flatten_channels(df)

# (it works on single vectors as well)
flatten_channels(df$age)
```

The value and missing reason channels of data frames with `interlaced` vectors
can similarly be accessed using the `value_channel()` and `na_channel()` helper
functions:

```{r}
value_channel(df)
na_channel(df)
```

## Next steps

So far, we've covered how interlacer's `read_interlaced_*` family
of functions enabled us to load `interlaced` columns that contain
separate challens for value and missing reasons. The `interlaced` type enables
us to create tidy and type-aware pipelines that can flexibly consider a
variable's value AND missing reasons.

In all the examples in this vignette, column types were automatically detected.
To explicitly specify value and missing column types, (and specify individual
missing reasons for specific columns), interlacer extends
`readr`'s `collector()` system. This will be covered in the next vignette,
`vignette("na-column-types")`.
