---
title: "Mutating Values and Missing Reasons"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mutating Values and Missing Reasons}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(dplyr)
library(interlacer)
```

When working with a "deinterlaced dataframe", care must be taken to ensure that
variables have a missing reason whenever a value is `NA`, and a 
value whenever a missing reason is `NA`. When this rule is violated, it creates
ambiguous states. For example if a variable has a values AND a missing reason,
it's not clear which one represents the "correct" state of the variable.
Similarly, if a variable is missing its value AND its missing reason, it's
probably a sign we made a mistake somewhere.

This means whenever we `mutate()` the values of a variable, the missing reasons
must also be updated, and vice versa. To illustrate this, let's load some
example data:

```{r}
(df <- read_interlaced_csv(
  interlacer_example("colors.csv"),
  na = c("REFUSED", "OMITTED", "N/A"),
))
```

Say we wanted to redact the values in the `age` variable, by setting their
missing reason to `REDACTED`:

```{r}
df |>
  mutate(
    .age. = "REDACTED"
  )
```

As you can see by the warning message, we've created an ambigous situation:
there are now rows where `age` and `.age.` both have values! We need to get rid
of all the `age` values:

```{r}
df |>
  mutate(
    .age. = "REDACTED",
    age = NA,
  )
```

Let's look at another example. Say our study was supposed to only let
participants choose between `RED` and `YELLOW` for their favorite colors --
but for some reason `BLUE` was included as an option because of a technical
glitch. In this situation, we'd want to set all responses that weren't `RED` and
`YELLOW` to be considered missing:

```{r}
df |>
  mutate(
    favorite_color = if_else(
      favorite_color %in% c("RED", "YELLOW"),
      favorite_color,
      NA
    )
  )
```

As you can see by the warning, we've created another invalid state, with some
`favorite_color` responses having neither values nor missing reasons. To fix
this, we need to make a corresponding mutation to the missing reason column:

```{r}
df |>
  mutate(
    favorite_color = if_else(
      favorite_color %in% c("RED", "YELLOW"),
      favorite_color,
      NA
    ),
    .favorite_color. = if_else(
      is.na(favorite_color) & is.na(.favorite_color.),
      "TECHNICAL_ERROR",
      .favorite_color.
    )
  )
```

To understand what's going on here, let's look at the mutation step by step:
First, where `favorite_color` is not `RED` or `YELLOW`, we set as a missing
value. In doing this, we've created a bunch of rows where both the value and
missing reason are absent. In next part of the mutation, we fill in the
`TECHNICAL_ERROR` missing reason for these rows into `.favorite_color.`,
resulting in a well-formed deinterlaced dataframe.

## An easier way with `coalesce_channels()`

As you can imagine, manually fixing the value & missing reason structure
of your dataframe for every mutation you do can get cumbersome! Luckily,
interlacer provides an easier way via `coalesce_channels()`:

```{r}
df |>
  mutate(
    .age. = "REDACTED",
  ) |>
  coalesce_channels(keep = "missing")

df |>
  mutate(
    favorite_color = if_else(
      favorite_color %in% c("RED", "YELLOW"),
      favorite_color,
      NA
    )
  ) |>
  coalesce_channels(default_reason = "TECHNICAL_ERROR")
```

`coalesce_channels()` should be run every time you mutate something in
a deinterlaced dataframe. It accepts two arguments `keep`, and `default_reason`.
With these paramters set, it fixes both possible problem cases as follows:

Case 1: BOTH a value and a missing reason exists

- Keep the value when `keep = 'value'`
- Keep the missing reason when `keep = 'missing'`

Case 2: NEITHER a value nor a missing reason exists

- Fill in the missing reason with `default_reason`

These rules allow us to mutate our deinterlaced variables without needing to
specify BOTH the values and missing reason actions -- we only need to think
about our operation one channel, and then a call to `coalesce_channels()`
takes care of the other.

## Creating New Columns

`coalesce_channels()` will also automatically create missing reason
columns if they don't automatically exist. This is useful for adding new
variables to your dataframe:

```{r}
df |>
  mutate(
    person_type = if_else(age < 18, "CHILD", "ADULT"),
    .after = person_id
  ) %>%
    coalesce_channels(default_reason = "AGE_UNAVAILABLE")
```

## Writing interlaced files

After you've made made changes to your data, you probably want to save them.
Interlacer provides the `write_interlaced_*` family of functions for this:

```{r, eval = FALSE}
write_interlaced_csv(df, "interlaced_output.csv")
```

This will combine the value and missing reasons into interlaced character
columns, and write the result as a csv. Alternatively, if you want to
re-interlace the columns without writing to a file for more control in the
writing process, you can use `interlace_channels()`:

```{r}
interlace_channels(df)
```

## Final note: Setting the global default reason

By default, `coalesce_channels()` will use `UNKNOWN_REASON` as the
default missing reason. Sometimes you want to use a different default value,
to act as the "catch-all" missing reason, so you don't have to constantly
specify it. To do this, set the global `default_missing_reason` option:

```{r}
options(default_missing_reason = -99)

tibble(
  a = c(1,2,3, NA, 5)
) |>
  coalesce_channels()
```
