---
title: "Other Approaches"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Other Approaches}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The [haven](https://haven.tidyverse.org/) package has functions for loading
native SPSS, SAS, and Stata native file formats into
special data frames that use column attributes and special values to keep track
of interlaces values and missing reasons. In this section I discuss the
advantages and disadvantages of these approaches, and how they compare with
interlacer. In the final section, I describe what I think would be an "ideal"
way to work with missingness in R, and the technical challenges we'd have to
overcome in order to implement it.

(Note: Future versions of interlacer could have the ability to convert haven
data frames to and from deinterlaced data frames, but I want to gauge interest
for this feature before I invest the time to implement it. If this is a
feature you'd use, [please let me know](mailto:kdh38@psu.edu)!)

## "Labelled" missing values

When SPSS *.sav files are loaded with haven via `haven::read_spss()`, values
and missing reasons are loaded into a single interlaced numeric vector:

```{r}
library(haven)

df_spss <- read_spss(
  interlacer_example("colors.sav")
)
```

## "Tagged" missing values

## The "ideal" approach

An "ideal" missing value API would make use of truly typed 
