---
title: "Other Approaches"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Other Approaches}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

interlacer was inspired by the [haven](https://haven.tidyverse.org/),
[labelled](https://larmarange.github.io/labelled/), and
[declared](https://dusadrian.github.io/declared/) packages. These packages
provide similar functionality to interlacer, but are more focused on
providing compatibility with missing reason data imported from SPSS, SAS, and
Stata. 

In this section I discuss some of the particularities of these 
approaches, and how they compare with interlacer.

(Note: Future versions of interlacer will have the ability to convert 
`haven_labelled` and `declared` types to and from `interlaced` types.)

## haven and labelled

The [haven](https://haven.tidyverse.org/) and
[labelled](https://larmarange.github.io/labelled/) packages rely on two
functions
for creating vectors that interlace values and missing reasons:
`haven::labelled_spss()` and `haven::tagged_na()`. Although they both create
`haven_labelled` vectors, they use very different methods for representing
missing values.

### "Labelled" missing values (`haven::labelled_spss()`)

When SPSS files are loaded with haven via `haven::read_spss()`, values
and missing reasons are loaded into a single interlaced numeric vector:

```{r}
library(interlacer, warn.conflicts = FALSE)
library(haven)
library(dplyr)

(df_spss <- read_spss(
  interlacer_example("colors.sav"), user_na = TRUE
))
```

Not just any numeric vector though, a `haven::labelled_spss()` numeric vector,
with attributes describing its value and missing value codes:

```{r}
attributes(df_spss$favorite_color)
```

These attributes adjust the behavior of functions like `is.na()`:

```{r}
is.na(df_spss$favorite_color)
```


This makes it easy to check if a value is a missing reason,
but you still have to filter out missing reasons before you do any aggregations:

```{r}
df_spss |>
  mutate(
    age_values = if_else(is.na(age), NA, age),
    favorite_color_missing_reasons = if_else(
      is.na(favorite_color), favorite_color, NA
    )
  ) |>
  summarize(
    mean_age = mean(age_values, na.rm = TRUE),
    n = n(),
    .by = favorite_color_missing_reasons
  )

df_spss |>
  mutate(
    age_next_year = if_else(is.na(age), NA, age + 1),
    .after = person_id
  )
```

It's a little bit of an improvement to working with raw coded values, because
you can use `is.na()`, and your codes get labels, so you don't have be
constantly looking up codes in your codebook. But it still falls short of 
interlacer's functionality for two key reasons:

Reason 1: With interlacer, your value column
can be whatever type you want: numeric, character, factor, etc. With labelled
missing reasons,
values and missing reasons need to be the same type, usually numeric codes.
This creates a lot more type gymnastics and potential errors when you're
manipulating them.

Reason 2: Even when the missing values are labelled in the `labelled_spss` type,
aggregations and other math operations are not protected. If you forget
to take out your missing values, you get incorrect results / corrupted data:

```{r}
df_spss |>
  mutate(
    favorite_color_missing_reasons = if_else(
      is.na(favorite_color), favorite_color, NA
    )
  ) |>
  summarize(
    mean_age = mean(age, na.rm = TRUE),
    n = n(),
    .by = favorite_color_missing_reasons
  )

df_spss |>
  mutate(
    age_next_year = age + 1,
    .after = person_id
  )
```


### "Tagged" missing values (`haven::tagged_na()`)

For loading Stata and SAS files, haven uses a "tagged missingness" approach to
mirror how these values are handled in Stata and SAS:

```{r}
(df_stata <- read_stata(
  interlacer_example("colors.dta")
))
```

This approach is deviously clever. It takes advantage of the way `NaN` floating
point values are stored in memory, to make it possible to have different
"flavors" of `NA` values. (For more info on how this is done, check out
[tagged_na.c](https://github.com/tidyverse/haven/blob/main/src/tagged_na.c) in
the source code for haven)

They still all act like regular `NA` values... but now they can include a single
character "tag" (usually a letter from a-z). This means that they work with
`is.na()` AND will not include missing reason codes in aggregations!

```{r}
is.na(df_stata$age)

mean(df_stata$age, na.rm = TRUE)
```

Unfortunately, you can't group by them, because `dplyr::group_by()` is not
tag-aware. :(

```{r}
df_stata |>
  mutate(
    favorite_color_missing_reasons = if_else(
      is.na(favorite_color), favorite_color, NA
    )
  ) |>
  summarize(
    mean_age = mean(age, na.rm = TRUE),
    n = n(),
    .by = favorite_color_missing_reasons
  )
```

Another limitation of this approach is that it requires values types to be
numeric, because the trick of "tagging" the `NA` values depends on the
peculiarities of how floating point values are stored in memory.

## declared

The [declared](https://dusadrian.github.io/declared/) package uses the
function`declared::declared()` for constructing interlaced vectors:

```{r}
library(declared)

(dcl <- declared(c(1, 2, 3, -99, -98), na_values = c(-99, -98)))
```

`declared` vectors are similar to `haven_labelled_spss` vectors, except with
a critical innovation: they store actual `NA` values where there are missing
values, and then keep track of the missing reasons entirely in the attributes
of the object:

```{r}
# All the missing reason info is tracked in the attributes
attributes(dcl)

# The data stored has actual NA values, so it works as you would expect
# with summary stats like `mean()`, etc.
attributes(dcl) <- NULL
dcl
```

This means aggregations work exactly as you would expect!

```{r}
dcl <- declared(c(1, 2, 3, -99, -98), na_values = c(-99, -98))

sum(dcl, na.rm = TRUE)
```

## interlacer

interlacer builds on the ideas of haven, labelled, and declared with following
goals:

### 1. Be fully generic: Add a missing value channel to *any* vector type

As mentioned above, `haven::labelled_spss()` only works with `numeric`
and `character` types, and `haven::tagged_na()` only works with `numeric` types. 
`declared::declared()` supports `numeric`, `character` and `date` types.

`interlaced` types, by contrast, can imbue *any* vector type with a missing value
channel:

```{r}
interlaced(list(TRUE, FALSE, "reason"), na = "reason")

interlaced(c("2020-01-01", "2020-01-02", "reason"), na = "reason") |>
  map_value_channel(as.Date)


interlaced(c("red", "green", "reason"), na = "reason") |>
  map_value_channel(factor)
```

Like `declared` vectors, the missing reasons are tracked in the attributes. But
unlike `declared`, missing reasons are stored as an entirely separate *channel*
rather than by tracking their indices:

```{r}
(int <- interlaced(c(1,2,3, -99, -98), na = c(-99, -98)))

attributes(int)

attributes(int) <- NULL
int
```

This data structure drives their functional API, described in (3) below.

### 2. Provide functions for reading / writing interlaced CSV files (not just SPSS
/ SAS / Stata files)

See `interlacer::read_interlaced_csv()`, etc.

### 3. Provide a functional API that integrates well into tidy pipelines

interlacer provides functions to facilitate working with the `interlaced` type
as a [Result type](https://en.wikipedia.org/wiki/Result_type),
a well-understood abstraction in functional programming. The functions `na()`
`map_value_channel()` and `map_na_channel()` all come from this influence.

The `na()` function creates an `interlaced` type by "lifting" a value into
the missing reason channel. This approach helps create a safer separation between
the value and missing reason channels, because it's always clear which channel
you're making comparisons on.

For example:

```{r}
# haven
labelled_spss(c(-99, 1, 2), na_values = -99) == 1 # value channel comparison
labelled_spss(c(-99, 1, 2), na_values = -99) == -99 # na channel comparison

# declared
declared(c(-99, 1, 2), na_values = -99) == 1 # value channel comparison
declared(c(-99, 1, 2), na_values = -99) == -99 # na channel comparison

# interlacer 
interlaced(c(-99, 1, 2), na = -99) == 1 # value channel comparison
interlaced(c(-99, 1, 2), na = -99) == na(-99) # na channel comparison
```

Similarly, `map_value_channel()` and `map_na_channel()` allow you to safely
mutate a particular channel, without touching the values of the
other channel. This interface is especially useful in tidy pipelines.

Finally, because the `interlaced` type is based on the `vctrs` type system, it
plays nicely with all the packages in the tidyverse.

## Questions for the future

### 1. More flexible missing reason channel types?

Earlier versions allowed arbitrary types to occupy
the missing reason channel (i.e. it was a fully generic Result<Value, Missing>
type). I ended
up constricting the missing reason channel to only allow `integer` or `factor`
types to help simplify the `na_cols()` specifications. When arbitrary types
are allowed, the `na_cols()` specs become quite long (e.g.
`column_name = factor(levels=c("REASON_1", "REASON_2")))`). As far as I can
tell, in 99.9% of the time, it is preferable to use `integer` and `factor`
missing reason channels over `double` and `character` ones, so for now I've
made the executive decision to only allow `integer` and `factor` types.

### 2. A better `na_cols()` specification?

Right now, missing values are supplied in a separate argument from
`col_types`. This means custom missing values get pretty far separated from
their `col_type` definitions:

```{r}
read_interlaced_csv(
  interlacer_example("stress.csv"),
  col_types = cols(
    person_id = col_integer(),
    current_stress = col_factor(
      levels = c("LOW", "MODERATE", "HIGH")
    ),
    time_management = col_factor(
      levels = c("POORLY", "FAIRLY_WELL", "WELL", "VERY_WELL")
    )
  ),
  na = na_cols(
    .default = c("REFUSED", "OMITTED", "N/A"),
    current_stress = c(.default, "DONT_KNOW", "DONT_UNDERSTAND"),
    time_management = c(.default, "NA_VACATION", "NA_OTHER")
  )
)
```

In an earlier version I created an extension of readr collectors, a family
of `icol_*` types, that allowed you to do something like this:

```{r, eval=FALSE}
read_interlaced_csv(
  interlacer_example("stress.csv"),
  col_types = cols(
    person_id = col_integer(),
    current_stress = icol_factor(
      levels = c("LOW", "MODERATE", "HIGH"),
      na = c("DONT_KNOW", "DONT_UNDERSTAND")
    ),
    time_management = col_factor(
      levels = c("POORLY", "FAIRLY_WELL", "WELL", "VERY_WELL"),
      na = c("NA_VACATION", "NA_OTHER")
    )
  ),
  na = c("REFUSED", "OMITTED", "N/A")
)
```

...I can't decide which interface I like better. Although the latter approach
feels cleaner because it folds custom missing reasons into the `cols`
definition, one disadvantage is that it cannot overwrite missing values (e.g.
I cannot set the missing reason on `person_id` to `NULL` as long as there's a
default missing reason specified).
It also feels a little "hackish" to extend readr's types in this way; I think
making use of the `na` parameter in my own `na_cols()` function provides me
with a little bit more insulation to changes from readr.

Anyway, if you have thoughts or opinions on any of these things, I'd really
[appreciate your feedback](mailto:kdh38@psu.edu)!
