---
title: "Other Approaches"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Other Approaches}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The [haven](https://haven.tidyverse.org/) package has functions for loading
native SPSS, SAS, and Stata native file formats into
special data frames that use column attributes and special values to keep track
of interlaces values and missing reasons. In this section I discuss the
advantages and disadvantages of these approaches, and how they compare with
interlacer. In the final section, I describe what I think would be an "ideal"
way to work with missingness in R, and the technical challenges we'd have to
overcome in order to implement it.

(Note: Future versions of interlacer could have the ability to convert haven
data frames to and from deinterlaced data frames, but I want to gauge interest
for this feature before I invest the time to implement it. If this is a
feature you'd use, [please let me know](mailto:kdh38@psu.edu)!)

## "Labelled" missing values

When SPSS *.sav files are loaded with haven via `haven::read_spss()`, values
and missing reasons are loaded into a single interlaced numeric vector:

```{r}
library(haven)

(df_spss <- read_spss(
  interlacer_example("colors.sav"), user_na = TRUE
))
```

Not just any numeric vector though, a `haven::labelled_spss()` numeric vector,
with attributes describing its value and missing value codes:

```{r}
attributes(df_spss$favorite_color)
```

These attributes adjust the behavior of functions like `is.na()`:

```{r}
is.na(df_spss$favorite_color)
```


So you still have to do the usual gymnastics in pipelines:

```{r}
df_spss |>
  mutate(
    age_values = if_else(is.na(age), NA, age),
    favorite_color_missing_reasons = if_else(
      is.na(favorite_color), favorite_color, NA
    )
  ) |>
  summarize(
    mean_age = mean(age_values, na.rm=T),
    n = n(),
    .by = favorite_color_missing_reasons
  )
```

It's a little bit of an improvement to working with raw coded values, because
you can use `is.na()`, and your codes get labels, so you don't have be
constantly looking up codes in your codebook. But it still falls short of the
interlacer approach for two key reasons.

Reason 1: With the interlacer approach
of having separate columns for values and missing reasons, your value column
can be whatever type you want: numeric, character, factor, etc. With labelled,
values and missing reasons need to be the same type, usually numeric codes.
This creates a lot more type gymnastics and potential errors when you're
manipulating them.

Reason 2: Keeping interlaced columns, even when the missing values are labelled,
means aggregations are not protected. If you forget to take out your missing
values, you get incorrect results:

```{r}
df_spss |>
  mutate(
    favorite_color_missing_reasons = if_else(
      is.na(favorite_color), favorite_color, NA
    )
  ) |>
  summarize(
    mean_age = mean(age, na.rm=T),
    n = n(),
    .by = favorite_color_missing_reasons
  )
```


## "Tagged" missing values

For loading Stata and SAS files, haven uses a "tagged missingness" approach to
mirror how these values are handled in Stata and SAS:

```{r}
(df_stata <- read_stata(
  interlacer_example("colors.dta")
))
```

This approach is deviously clever. It takes advantage of the way `NaN` floating
point values are stored in memory, to make it possible to have different
"flavors" of `NA` values. (For more info on how this is done, check out
[tagged_na.c](https://github.com/tidyverse/haven/blob/main/src/tagged_na.c) in
the source code for haven)

They still all act like regular `NA` values... but now they can include a single
character "tag" (usually a letter from a-z). This means that they work with
`is.na()` AND will not include missing reason codes in aggregations!

```{r}
is.na(df_stata$age)

mean(df_stata$age, na.rm=TRUE)
```

Unfortunately, you can't group by them, because `dplyr::group_by()` is not
missing tag-aware :(

```{r}
df_stata |>
  mutate(
    favorite_color_missing_reasons = if_else(
      is.na(favorite_color), favorite_color, NA
    )
  ) |>
  summarize(
    mean_age = mean(age, na.rm=T),
    n = n(),
    .by = favorite_color_missing_reasons
  )
```

Another limitation of this approach is that it requires values types to be
numeric, because the trick of "tagging" the `NA` values depends on the
peculiarities of how floating point values are stored in memory. Again,
keeping separate columns for values and missing reasons solves all these issues.

## The "ideal" approach

The biggest downside of keeping separate columns for values and missing reasons
are the invalid states that come up when you start trying to mutate your data
frames. `coalesce_channels()` helps a lot, but it's not ideal.

I think the ideal way to handle missing reasons would be to implement a proper
generic [`Result` type](https://en.wikipedia.org/wiki/Result_type) natively
into R's type system. A real `Result` type would act similar to haven's
`haven::tagged_na()`, but be a container for any type of value, not only
missing values.

In an early attempt of this library, I tried using nested data frames for this
effect:

```{r}
df_interlaced <- read_interlaced_csv(
  interlacer_example("colors.csv"),
  na = c("REFUSED", "OMITTED", "N/A")
)

(df_nested <- tibble(
  person_id = tibble(
    v = df_interlaced$person_id,
    m = df_interlaced$.person_id.,
  ),
  age = tibble(
    v = df_interlaced$age,
    m = df_interlaced$.age.,
  ),
  favorite_color = tibble(
    v = df_interlaced$favorite_color,
    m = df_interlaced$.favorite_color.,
  )
))
```

This sort of works, because we can use `$v` and `$m` to reference separate
channels of the data frame. Unfortunately it requires creating separate columns
when grouping:

```{r}
df_nested |>
  mutate(
    favorite_color_missing = favorite_color$m
  ) |> 
  summarize(
    mean_age = mean(age$v, na.rm=T),
    n = n(),
    .by = favorite_color_missing
  )
```

And mutations get ugly... (There's probably a way to tickle this back to a
nicely displayed data frame, but I couldn't find it)

```{r}
df_nested |>
  mutate(
    favorite_color = if_else(
      favorite_color$v %in% c("RED", "YELLOW"),
      tibble(v = favorite_color$v, m = NA),
      tibble(v = NA, m = "TECHNICAL_ERROR")
    )
  )
```

If we were to implement this somehow as a custom native type in R, I'd want
syntax something like this instead:

```{r, eval = FALSE}
df_mutated <- df |>
  mutate(
    favorite_color = if_else(
      favorite_color %in% c("RED", "YELLOW"),
      favorite_color,
      missing_reason("TECHNICAL_ERROR")
    )
  
df_mutated |>
  summarize(
    mean_age = mean(age, na.rm=T),
    n = n(),
    .by = missing_reason(favorite_color)
  )
```

This would be "ideal" in my book: we can use values as usual, but anytime
we want to access the "missing reason" channel, we can wrap it in a
`missing_reason()` (similar to how `haven::tagged_na()` works). It's type safe
and super ergonomic. But implementing this would be a major headache and
involve very intimate knowledge of R internals... (@Hadley Wickham if by some
miracle you're reading this, could we talk sometime??)

So this is why I'm using the present current "deinterlaced data frame"
approach. It is easy to understand and use, even though it's not "perfect"
from a strongly typed functional programming perspective. If there's enough
demand for missing-reason-aware tooling in R though, it might convince me
to go down the "generic tagged type" rabbit hole...
[Please drop me a line](mailto:kdh38@psu.edu) to let me know what you think!
